Autonomous Oceanographic Sampling Designs using Excursion Sets for Multivariate Gaussian random fields

### 2020-06-05 main_new.tex
Jo incorporated some changes in main file.

There are several challenges with sampling with autonomous platforms.
\kc{You seem to jump the gun in stating a challenge without
  connecting it to the sampling problem. Introduce a sentence as to
  why models (as a proxy) are important for \emph{understanding} ocean
  processes} First, full numerical ocean models based on complex
differential equations cannot be run onboard a marine robot with
limited computational capacity.
% provide accurate results for online
% operations because the onboard computers lack the computational power
% required.
Hence statistical proxy models of the environment must be
used. For our purpose, we rely on Gaussian process (GP)
representations of the ocean variables of interest.\kc{Sorry....but
  this makes no sense whatsoever. What is the connection of you using
  a GP with sampling? You need to make a case first as to the
  viability of using GP in a marine robot first as a potential viable
  option, before you state why you want to use a GP}. Second, with
undersampling issues \kc{what issues? If there are issues, you state
  them first and then suggest solutions.} one must usually plan for
active learning during the operation, where a relevant criterion is
used to extract the most valuable designs. The acquired in-situ
information must then be assimilated into the statistical model
onboard the platform, so that it can be used to re-compute the
sampling criterion and in this way inform decisions on where to sample
next. We use a criterion focusing on the prediction of excursion sets
(ESs) for the multivariate spatial processes of interest. Third, the
number of sampling designs is enormous, creating a trade-off between
optimization and computability. One must often compromise in both
development and practical operations, and we will rely on
short-horizon optimization routines for sequential designs among
nearest neighbors points in a spatial grid (waypoint graph).

@@

With this in mind, we have designed AUV sampling strategies for
reducing the spatially-integrated Bernoulli variance (IBV) in the ES
indicator variables. 
ES of random fields are thoroughly presented in \citep{adler2009random}. Recent work on ESs and associated excursion
probabilities (EPs) connected to spatial statistics include
\cite{french2013spatio,bolin2015excursion,french2016credible}. Our
focus is on uncertainty characterization and how this is influenced by
sampling. In this sense, our work is similar to
\cite{bect2012,chevalier2014fast,azzimonti2016quantifying} who
describe analytical results for the IBV for univariate processes.

@@

@@

Traditional data collection to understand such phenomena at sea has
typically been based on static buoys, floats, or ship-based methods,
with significant logistical limitations that directly impact coverage
and sampling resolution. Modern methods using satellite remote-sensing
provide large-scale coverage but have limited resolution, are limited
to sensing the surface, and are impacted by cloud cover. Numerical
ocean models similarly find it challenging to provide detail at fine
scale \citep{Lermusiaux:2006}. The advent of mobile robotic platforms
\citep{Bellingham07} has resulted in significant contributions to
environmental monitoring and sampling. In particular, AUVs have
advanced the state of in-situ sampling, and made robotics an integral
part of ocean observation, filling parts of the undersampling gap
\cite{rudnick03,rudnick18}(See Fig. \ref{fig:envir1}).

@@

Adaptive in-situ AUV sampling of an evolving frontal feature has been
explored in \cite{fronts11,Zhang2012,Pinto2018,costa19}. These
approaches typically use a reactive-adaptive scheme, whereby
exploration does not rely on a statistical model of the environment,
but rather adaptation is based on closing the sensing and actuation
loop. Myopic sampling, i.e. stage-wise selection of the path (on the
waypoint graph), has been used for surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or entropy. These criteria are widely adopted in
the statistics literature on spatio-temporal design as well
\cite{bueso1998state,zidek2019monitoring}, but variance and entropy
reduction are independent of the actual data realizations under the
assumptions of GP models, so limited decisional flexibility is created
for sequential designs \kc{should this be, ``...so it has limited
  decisional flexibility''?}. The use of data-driven adaptive criteria
was introduced to include more targeted sampling of regions of
scientific interest in \cite{Low2009} and \cite{fossuminformation}. In
this paper, we focus on mapping the river plume by rewarding designs
that improve the classification of ES in temperature and salinity.

@@

The definitions of the sets of interest are introduced in Section \ref{sec:ES}, while the closed-form expressions for the expected uncertainty reduction are derived in Section \ref{sec:EIBV}. Section \ref{Sec:UnivarEx} provides an instructive bivariate example relevant for sampling in the temperature and salinity application.

@@

In order to formalize the problem in mathematical terms, let us
consider a multivariate (i.e., vector-valued) Gaussian random field
$(\gp[\x])_{\x \in \mathcal{M}}$, where $\mathcal{M}$ is an arbitrary
set (typically in geostatistical studies, a geographical subset of interest) and the
dimensionality of the output $\no \geq 1$ is arbitrary \kc{Is the
  definition of such a 'p' and its dimensionality, obvious? This to me
  springs out of nowhere.}. Assume furthermore that the set of interest
is characterized in terms of a subset of values of the variables, say
$\T \subset \R^{\no}$, and denote the corresponding set of index
points by
$$
\es=\gp^{-1}(\T)=\{\x \in \mathcal{M}: \gp[\x] \in \T\}.
$$
Our goal here is to develop approaches to quantify and improve the
characterization of uncertainties on such sets \kc{$\es$?} in the
context of multivariate random fields.

@@

We let $d \in \mathcal{D}$ denote a concatenation of sampling designs with $q$ sampling locations $\bu_d=\left(u_d^{(1)},\ldots, u_d^{(q)}\right)$. The
measurement model is defined by 
\begin{equation}\label{eq:meas}
    \bY_d=G_d \bz(\bu_d) + \epsilon, \hspace{3mm} \epsilon \sim N(0,R_d).
\end{equation}
Here, $\bz(\bu_d)$ is the size $pq$ vector of GP variables at the
sampling locations, $G_d$ is the design matrix which in principle
allows for observing subsets of the vector-valued field at some of the
$q$ locations (i.e. the matrix has fewer rows than columns). If the
additive error has independent elements, the covariance matrix $R_d$
is diagonal. In our setting temperature and salinity measurements are
extracted from different sensors, so the independence assumption
between variables is not unrealistic, but this is not necessarily the
case for other kinds of measurements on the AUV. It is further common
in geostatistics to assume spatially independent measurement errors,
and we also do so. But our methodology for constructing sampling
designs does not rely on these independence assumptions.

@@

Assuming now that $q$ observations are available, one wishes to
predict $\gp[\x]$. The multivariate Kriging equations holds with size
$p$ vector of conditional means:  
\begin{equation}\label{eq:cokrig}
\bbm(\x)=\bmu(\x)+\lambda(\x,\bu_d)^T \left( \by_d-G_d \bmu(\bu_d)\right),
\end{equation}
where $\bmu$ is size $p$ unconditional mean function of $Z$, $\by_d$
is the realized vector of observed responses, and
$\lambda(\x,\bu_d)=K^{-1}_y(\bu_d,\bu_d) k(\bu_d, \bu)$. Here,
the data covariance is
$K_y=\mbox{Cov}(\bY_d,\bY_d)=G_d K(\bu_d,\bu_d)G_d^T+R_d$ while
the cross-covariance is $\mbox{Cov}(\bY_d,Z(\x))=G_d k(\bu_d, \x)$, assuming the measurement
error is independent of the random fields.  The Kriging (cross-)covariance function can also be expressed in the same vein via
a $p \times p$ matrix 
\begin{equation}\label{eq:krig:cov}
\Sigma(\x,\x')=K(\x,\x')-\lambda(\x,\bu_d)^T K_y(\bu_d, \bu_d) \lambda(\x,\bu_d).
\end{equation}

@@

% In this section, 
We calculate EPs and the Bernoulli variance for a few bivariate
Gaussian distributions to illustrate the concepts articulated above.
Fig. \ref{illus_bivarDens} shows contour plots of three different
densities with increasing correlation $\gamma$ between temperature and
salinity. \kc{What is the source of this data and this figure,
  i.e. how did you derive this plot?}
\begin{figure}[h!] \centering
  \includegraphics[width=0.99\textwidth]{Figures/illus_bivar.pdf}
  \caption{Density contour plots with different correlations between
    temperature and salinity. The densities have unit variance and the
    thresholds are identical to the mean values $5^o C$ and
    $30 mg/l$. X-axis is temperature and y-axis is salinity.}
\label{illus_bivarDens}
\end{figure}
Here, the thresholds are set equal to the mean $\mu_1=t_1=5^o C$, and
salinity $\mu_2=t_2=30$ mg/l, and the EP is $p=P(Z_1 < 5,Z_2 < 30)$.
The displayed densities have unit standard deviations for both
temperature and salinity, but we also study the effect of doubling the
standard deviations.

Table \ref{tab:sim_rhoab} shows the initial EPs and the associated
Bernoulli variance (second row) for the examples indicated in Fig.
\ref{illus_bivarDens}. The EPs increase with the correlation as there
is a strong tendency to have concurrently low temperature and salinity
\kc{Same question as above. What is the basis for this correlation? Is
  this just an example or derived from actual data in the mixed layer
  in the front?}. The Bernoulli variance is similarly large for high
correlations. EPs and Bernoulli variances are the same for standard
deviation $1$ or $2$, which implies that high variability in
temperature and salinity is not captured in the $p(1-p)$ expression.

@@

Table \ref{tab:sim_rhoab} (bottom two rows) shows results of expected
Bernoulli variance calculations. This is presented for a design
gathering both data types, and for a design with temperature
measurements alone. Both data are modeled by
$(Y_1,Y_2)^t=(Z_1,Z_2)^t+N(0,0.5^2I_2)$, while $Y_2=Z_2+N(0,0.5^2)$
when only temperature is measured \kc{I find this notation somewhat
  confusing. Earlier you had set up t1 and t2 to denote temp and
  salinity respectively. It might help to have \emph{some} notational
  similarity with that; otherwise it becomes mind bending I believe.}.
For this illustration, Table \ref{tab:sim_rhoab} shows that the
expected Bernoulli variance gets lower with larger standard deviations
$\sigma_1$ and $\sigma_2$ (right columns). The reduction of Bernoulli
variance is largest for the cases with high correlation
$\gamma$. Albeit smaller, there is also uncertainty reduction when
only temperature is measured (bottom row), especially when temperature
and salinity are highly correlated. When correlation is low
($\gamma=0.2$), there is little information about salinity in the
temperature data, and therefore less uncertainty reduction. In an
application with fresh cold water from a river source, the temperature
and salinity variables will not only be interdependent, but will also
likely show dependence in the spatial dimension. This in turn will
impact the design criteria when we evaluate the information measure by
integrating over several locations (Section \ref{sec:simulations}).

\kc{I feel this section above, which was present in the earlier
  submitted version too, but to which I did not comment, is
  significant as an example to the reader. But, as noted above, I
  don't understand the provenance of the data. It isn't made up,
  right?}

@@

Let $d^{j,s}$ denote design number $j$ at stage $s$ of a sequential
survey. If this design is selected, data $\by_{d^{j,s}}$ will be
gathered.  We further let
$\mathcal{Y}^{s-1} =\{(d^{j,1}, \by_{d^{j,1}}),\ldots, (d^{j,s},
\by_{d^{j,s}}) \}$ denote the set of all data gathered until stage
$s-1$ according to selected design $d^{j,1},\ldots,d^{j,s}$.  This
defines the AUV position at the stage $s$.  The data further defines
the conditional GP with mean
$\bbm^{s-1}(\x)=E(\gp[\x]|\mathcal{Y}^{s-1})$ from
Eq. (\ref{eq:cokrig}) and (cross) covariance matrix
$\Sigma^{s-1}=\mbox{Var}(\gp[\x]|\mathcal{Y}^{s-1})$ from
Eq. (\ref{eq:krig:cov}).

At the next stage, we choose where measurements $\by_{d^{j,s}}$ should
be gathered with design $d^{j,s}$. For the triangularized waypoint
graph that we use in our application, there are six possible
directions for $j$, and we must evaluate the design criterion for all
these.  The goal of the sampling design is to minimize the expected
uncertainty of random sets, and specifically the EIBV for our
application. We hence focus on choosing designs for data modeled by
equation \eqref{eq:meas}, to conduct efficient sequential uncertainty
reduction according to the EIBV criterion.  Notably, compared with the
static design $d$ in equation \eqref{eq:eibv}, the optimal sequential
design not only considers the best current design grid point, but also
how the data gathered at this node could inform future sampling at
successive stages.

@@

This strategy does not account for the uncertainty in the GP
variables, but is instead applicable only if one design (node) has EPs
closer to $0.5$ (see Table \ref{tab:sim_rhoab}, line \kc{row?}
two). Additionally, this strategy does not account for spatial
correlation. This strategy lacks memory of where it has been and where
the uncertainty has been reduced, and therefore susceptible to local
minima.

@@

Note that this strategy gives a sequential conditional version of
expression (\ref{eq:eibv}). Now $\mathcal{Y}^{s-1}$ is available, and
the expectation is conditional to this data. A similar closed-form
calculation for EIBV is hence applicable in Eq.  \eqref{critSEQ}
\kc{Strangely in my TeX compilation this points to the earlier
  definition of $d^{*,s}$. Seems like some index is messed up in the
  sources; all indexes should perhaps be checked?} using the updated
GP model from stage $s-1$. Once the data is collected for the best
\kc{??}  design at stage $s$, the GP model is updated again. The mean
$\bbm^{s}$ and covariance matrix $\Sigma^{s}$ are used to compute the
next design at stage $s+1$, and so on.

@@

We now extend the myopic strategy to a look-ahead strategy, which is
optimal when one can gather only two more stages of
measurements. \kc{How can you be sure of this?} In addition to the
next stage of measurements $\by^{j,s}$, this look-ahead strategy,
anticipates the subsequent design indexed $j_2$ with data
$\by^{j_2,s+1}$ when choosing the current design $d^{j,s}$.  The
selected design is:
\begin{eqnarray}\label{critLA}
    d^{*,s} &=& \mbox{argmin}_{j} \left\{ E_{\by_{d^{j,s}}|\mathcal{Y}^{s-1}} \left( \mbox{argmin}_{j_2}  V_{j_2} \right) \right\}, \\
V_{j_2} & = & \int_{\mathcal{M}} E_{\by_{d^{j_2,s+1}}|\mathcal{Y}^{j,s}} \left( p(\x;\mathcal{Y}^{j,s+1}) ( 1-p(\x;\mathcal{Y}^{j,s+1}) ) \right) d\mes(u), \nonumber \\
    p(\x;\mathcal{Y}^{j,s+1}) &=& P(Z(\x) \in T |\by_{d^{j_2,s+1}},\mathcal{Y}^{j,s}). \nonumber
\end{eqnarray}
Here, $\mathcal{Y}^{j,s}=\{\by_{d^{j,s}},\mathcal{Y}^{s-1}\}$ and
$\mathcal{Y}^{j,s+1}=\{\by_{d^{j_2,s+1}},\mathcal{Y}^{j,s}\}$
represent the sets of data variables, and the expectations are
calculated with respect to the conditional densities given data at
that stage, implying that we have two levels of intermixed
minimizations and integrals.  We solve the first expectation by Monte
Carlo sampling of data $\by_{d^{j,s}}$ from its conditional
distribution. For each of these data samples, the second expectation
is solved using the closed-form expressions for EIBV from Section
\ref{sec:ESEP}.

Even though the strategy looks at two sampling stages, it is only used to find
the current best design. When data are collected, the GP model is
updated, and the mean $\bbm^{s}$ and covariance matrix $\Sigma^{s}$ are
used to compute the next design at stage $s+1$, now anticipating what
stage $s+2$ could offer, and so on.
This look-ahead approach is much more computationally demanding than
the myopic strategy, so for practical implementation, we prune paths
in the evaluation of Eq. \eqref{critLA}. This means that we use the myopic
strategy to rank the three best designs following the first stage alone, and
for each of these preferred designs, we undertake the look-ahead calculations.

>>>>

The
advent of robust mobile robotic platforms \citep{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling in the ocean. In particular, autonomous underwater vehicles (AUVs), have
advanced the state of sampling and consequently have made robotics an
integral part of ocean observation; our previous work has contributed
to this effort \citep{das11b,Das2015,fossuminformation,fossum18b}.

The potential that lies in using information-based sampling methods, which we
highlight in this paper, is to do efficient environmental sampling and
overcome common challenges as discussed next. First, full numerical ocean models based on complex
differential equations cannot be run onboard a marine robot with
limited computational capacity.
% provide accurate results for online
% operations because the onboard computers lack the computational power
% required.
Hence statistical proxy models of the environment must be used. For our purpose, we rely on Gaussian process (GP)
representations of the ocean variables of interest because they are computationally convenient and yet provide enough flexibility to realistically model the spatial variability and dependence, and the correlation between multiple processes. Second, because of limited resources to sample the large oceanographic domains, one must plan for
active learning during the operation, where a relevant criterion is
used to extract the most valuable designs. The acquired in-situ
information must then be assimilated into the statistical model
onboard the platform, so that it can be used to re-compute the
sampling criterion and in this way inform decisions on where to sample
next. We use a criterion focusing on the prediction of excursion sets
(ESs) for the multivariate spatial processes of interest. Third, the
number of sampling designs is enormous, creating a trade-off between
optimization and computability. One must often compromise in both
development and practical operations, and we will rely on
short-horizon optimization routines for sequential designs among
nearest neighbors points in a spatial grid (waypoint graph).

@@

With this in mind, we design AUV sampling strategies for reducing the spatially-integrated Bernoulli variance (IBV) in the ES
indicator variables. 
ES of random fields are thoroughly presented in \citep{adler2009random}. Recent work on ESs and associated excursion
probabilities (EPs) connected to spatial statistics include
\cite{french2013spatio,bolin2015excursion,french2016credible}. Our
focus is on uncertainty characterization and how this is influenced by
sampling. In this sense, our work is similar to
\cite{bect2012,chevalier2014fast,azzimonti2016quantifying} who
describe analytical results for the IBV for univariate processes.

@@

Other statistical
work in the oceanographic domain include \cite{wikle2013modern}
focusing on hierarchical statistical models; \cite{sahu2008space},
studying spatio-temporal models for sea surface temperature and
salinity data; and \cite{mellucci2018oceanic} looking at the
statistical prediction of features using an underwater glider.
In this paper the focus is not on statistical modeling, but rather on statistical results and computations for efficient sampling designs.

@@

Traditional data collection to understand such phenomena at sea has
typically been based on static buoys, floats, or ship-based methods,
with significant logistical limitations that directly impact coverage
and sampling resolution. Modern methods using satellite remote-sensing
provide large-scale coverage but have limited resolution, are limited
to sensing the surface, and are impacted by cloud cover. Numerical
ocean models similarly find it challenging to provide detail at fine
scale \citep{Lermusiaux:2006}. Mobile robotic platforms
\citep{Bellingham07} have contributed significantly to environmental monitoring and sampling. In particular, AUVs have
advanced the state of in-situ sampling, and made robotics an integral
part of ocean observation, filling parts of the undersampling gap
\cite{rudnick03,rudnick18}(See Fig. \ref{fig:envir1}).

@@

Adaptive in-situ AUV sampling of an evolving frontal feature has been
explored in \cite{fronts11,Zhang2012,Pinto2018,costa19}. These
approaches typically use a reactive-adaptive scheme, whereby
exploration does not rely on a statistical model of the environment,
but rather adaptation is based on closing the sensing and actuation
loop. Myopic sampling, i.e. stage-wise selection of the path (on the
waypoint graph), has been used for surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or entropy. These criteria are widely adopted in
the statistics literature on spatio-temporal design as well
\cite{bueso1998state,zidek2019monitoring}, but variance and entropy
reduction are independent of the actual data realizations under the
assumptions of GP models, so it has limited
decisional flexibility. The use of data-driven adaptive criteria
was introduced to include more targeted sampling of regions of
scientific interest in \cite{Low2009} and \cite{fossuminformation}. In
this paper, we focus on mapping the river plume by rewarding designs
that improve the classification of ES in temperature and salinity.

@@

We define the random sets of interest in Section \ref{sec:ES}, while the closed-form expressions for the expected uncertainty reduction are derived in Section \ref{sec:EIBV}. Section \ref{Sec:UnivarEx} provides an instructive bivariate example relevant for sampling in the temperature and salinity application.

@@

In order to formalize the problem in mathematical terms, let us
consider a multivariate (i.e., vector-valued) Gaussian random field
$(\gp[\x])_{\x \in \mathcal{M}}$, where $\mathcal{M}$ is an arbitrary
set (typically in geostatistical studies, a geographical subset of interest) and the
dimensionality of the output $\no \geq 1$ is arbitrary. We have $\no = 2$ in the example with temperature and salinity in the river plume.

Assume furthermore that the set of interest
is characterized in terms of a subset of values of the variables, say
$\T \subset \R^{\no}$, and denote the corresponding set of index
points by
$$
\es=\gp^{-1}(\T)=\{\x \in \mathcal{M}: \gp[\x] \in \T\}.
$$
Our goal here is to develop approaches to quantify and improve the
characterization of uncertainties on such sets $\es$, in the
context of multivariate random fields.

@@

We denote a design by $d \in \mathcal{D}$, where the concatenation of designs with $q$ sampling locations $\bu_d=\left(u_d^{(1)},\ldots, u_d^{(q)}\right)$ is $\mathcal{D}$. The
measurement model is defined by 
\begin{equation}\label{eq:meas}
    \bY_d=G_d \bz(\bu_d) + \epsilon, \hspace{3mm} \epsilon \sim N(0,R_d).
\end{equation}
Here, $\bz(\bu_d)$ is the size $pq$ vector of GP variables at the
sampling locations, and $G_d$ is the design matrix which in principle
allows for observing subsets of the vector-valued field at some of the
$q$ locations (i.e. the matrix has fewer rows than columns). If the
additive error has independent elements, the covariance matrix $R_d$
is diagonal. In our setting temperature and salinity measurements are
extracted from different sensors, so the independence assumption
between variables is not unrealistic, but this is not necessarily the
case for other kinds of measurements on the AUV. It is further common
in geostatistics to assume spatially independent measurement errors,
and we also do so. But our methodology for constructing sampling
designs does not rely on these independence assumptions.

@@

Assuming now that $q$ observations are available, one wishes to
predict $\gp[\x]$. The multivariate Kriging equations holds with size
$p$ vector of conditional means:  
\begin{equation}\label{eq:cokrig}
\bbm(\x)=\bmu(\x)+\lambda(\x,\bu_d)^T \left( \by_d-G_d \bmu(\bu_d)\right),
\end{equation}
where $\bmu$ is size $p$ unconditional mean function of $Z$, $\by_d$
is the realized vector of observed responses, and
$\lambda(\x,\bu_d)=K^{-1}_y k(\bu_d, \bu)$. Here,
the data covariance is
$K_y=\mbox{Cov}(\bY_d,\bY_d)=G_d K(\bu_d,\bu_d)G_d^T+R_d$ while
the cross-covariance is $\mbox{Cov}(\bY_d,Z(\x))=G_d k(\bu_d, \x)$, assuming the measurement
error is independent of the random fields.  The Kriging (cross-)covariance function can also be expressed in the same vein via
a $p \times p$ matrix 
\begin{equation}\label{eq:krig:cov}
\Sigma(\x,\x')=K(\x,\x')-\lambda(\x,\bu_d)^T K_y(\bu_d, \bu_d) \lambda(\x,\bu_d).
\end{equation}

@@

% In this section, 
We calculate EPs and the Bernoulli variance for a few bivariate
Gaussian distributions for temperature and salinity to illustrate the concepts articulated above. The parameter inputs are made up, but they are in a realistic range as we have in the simulation study and real-world example below. The thresholds are set equal to the mean; $\mu_1=t_1=5^o C$ for temperature and  $\mu_2=t_2=30$ mg/l for salinity, and we play with the temperature and salinity correlation and variances to study the effect on the EP and expected Bernoulli variance.

Fig. \ref{illus_bivarDens} shows contour plots of three different
densities with increasing correlation $\gamma$ between temperature and
salinity. 
\begin{figure}[h!] \centering
  \includegraphics[width=0.99\textwidth]{Figures/illus_bivar.pdf}
  \caption{Density contour plots with different correlations between
    temperature and salinity. The densities have unit variance and the
    thresholds are identical to the mean values $5^o C$ and
    $30 mg/l$. X-axis is temperature and y-axis is salinity.}
\label{illus_bivarDens}
\end{figure}
The displayed densities have unit standard deviations for both
temperature and salinity, but we also study the effect of doubling the
standard deviations.

Table \ref{tab:sim_rhoab} shows the initial EPs and the associated
Bernoulli variance (second row) for the examples indicated in Fig.
\ref{illus_bivarDens}. The EPs increase with the correlation as there
is a strong tendency to have concurrently low temperature and salinity. The Bernoulli variance is similarly large for high
correlations. EPs and Bernoulli variances are the same for standard
deviation $1$ or $2$, which implies that high variability in
temperature and salinity is not captured in the $p(1-p)$ expression.

@@

Table \ref{tab:sim_rhoab} (bottom two rows) shows results of expected
Bernoulli variance calculations. This is presented for a design
gathering both data types, and for a design with temperature
measurements alone. When both data are gathered, the measurement model is
$(Y_{d,1},Y_{d,2})^t=(Z_1,Z_2)^t+\bepsilon$, with $\bepsilon \sim N(0,0.5^2I_2)$, while $Y_d=Z_1+\epsilon$, $\epsilon \sim N(0,0.5^2)$ when only temperature is measured.
For this illustration, Table \ref{tab:sim_rhoab} shows that the
expected Bernoulli variance gets lower with larger standard deviations
$\sigma_1$ and $\sigma_2$ (right columns). The reduction of Bernoulli
variance is largest for the cases with high correlation
$\gamma$. Albeit smaller, there is also uncertainty reduction when
only temperature is measured (bottom row), especially when temperature
and salinity are highly correlated. When correlation is low
($\gamma=0.2$), there is little information about salinity in the
temperature data, and therefore less uncertainty reduction. In an
application with fresh cold water from a river source, the temperature
and salinity variables will not only be interdependent, but will also
likely show dependence in the spatial dimension. This in turn will
impact the design criteria when we evaluate the information measure by
integrating over several locations (Section \ref{sec:simulations}).

@@

Let $d^{j,s}$ denote design number $j$ at stage $s$ of a sequential
survey. For the AUV sampling situation, the choice of design entails a commitment to one of many possible navigation directions on the waypoint graph. Once the choice has been made, the AUV steers in the chosen direction and this entails sampling locations $\bu_{d^{j,s}}$ and data $\by_{d^{j,s}}$ is gathered.  We let
$\mathcal{Y}^{s-1} =\{(d^{*,1}, \by_{d^{*,1}}),\ldots, (d^{*,s-1},
\by_{d^{*,s-1}}) \}$ denote the set of all data gathered until stage
$s-1$ according to selected design $d^{*,1},\ldots,d^{*,s-1}$.

The design and data define
the conditional GP with mean
$E(\gp[\x]|\mathcal{Y}^{s-1})$ and (cross) covariance matrix
$\mbox{Var}(\gp[\x]|\mathcal{Y}^{s-1})$. These are efficiently updated in a sequential manner keeping track of the mean and covariance kernel of the GP over the stages. As extensions of Eq. \eqref{eq:cokrig} and \eqref{eq:krig:cov}, we have
\begin{eqnarray}\label{eq:krigSUR}
\mu^{s}(\x) &=&\mu^{s-1}(\x)+\lambda^{[s,s]}(\x,\bu_{d^{j,s}})^T (\by_{d^{j,s}}-G_{d^{j,s}} \mu(\x)), \nonumber \\
k^{s}(\x,\x') &=& k^{s-1}(\x,\x')-\lambda^{[s,s]}(\x,\bu_{d^{j,s}})^T K_{Y_{d^{j,s}}} \lambda^{[s,s]}(\x',\bu_{d^{j,s}}),
\end{eqnarray}
where $\lambda^{[s,s]}(\x,\bu_{d^{j,s}})=K^{-1}_{Y_{d^{j,s}}} k(\bu_{d^{j,s}},\x)$ and the data covariance is $K^{-1}_{Y_{d^{j,s}}}=G_{d^{j,s}} K^s(\bu_{d^{j,s}},\bu_{d^{j,s}})G^T_{d^{j,s}}+R_{d^{j,s}}$, with measurement model and design matrix defined as in equation \eqref{eq:meas}

At stage $s$, the robot must then choose design $d^{j,s}$ where measurements $\by_{d^{j,s}}$ should
be gathered. For the triangularized waypoint
graph all six possible
directions must be evaluated to minimize the expected uncertainty of random sets, and in particular the EIBV for our 
application. We hence focus on choosing designs for data modeled by
equation \eqref{eq:meas}, to conduct efficient sequential uncertainty
reduction according to the EIBV criterion.  Notably, compared with the
static design $d$ in equation \eqref{eq:eibv}, the optimal sequential
design not only considers the best current design grid point, but also
how the data gathered at this node could inform future sampling at
successive stages.

@@

This strategy does not account for the uncertainty in the GP
variables, but is instead applicable only if one design (node) has EPs
closer to $0.5$ (see Table \ref{tab:sim_rhoab}, row
two). Additionally, this strategy does not account for spatial
correlation. This strategy lacks memory of where it has been and where
the uncertainty has been reduced, and therefore susceptible to local
minima.

@@

Note that this strategy gives a sequential conditional version of
expression (\ref{eq:eibv}). Now $\mathcal{Y}^{s-1}$ is available, and
the expectation is conditional to this data. A similar closed-form
calculation for EIBV is hence applicable in Eq.  \eqref{critSEQ} using the updated
GP model from stage $s-1$ as defined via Eq. \eqref{eq:krigSUR} in the results from Proposition \ref{propo2}. Once the data are collected for the best design $d^{*s}$ at stage $s$, the GP model is updated again. The mean and covariance are used to compute the next design at stage $s+1$, and so on.

@@

We now extend the myopic strategy to a look-ahead strategy. We consider two stages of
measurements, and the strategy is then optimal in the sense that it accounts consistently for the expectations and minimizations in these two stages, but still without including any planning beyond those two steps. In addition to the
next stage of measurements $\by^{j,s}$, this look-ahead strategy anticipates the subsequent design indexed $j_2$ with data
$\by^{j_2,s+1}$ when choosing the current design $d^{j,s}$. The
selected design is:
\begin{eqnarray}\label{critLA}
    d^{*,s} &=& \mbox{argmin}_{j} \left\{ E_{\by_{d^{j,s}}|\mathcal{Y}^{s-1}} \left( \mbox{argmin}_{j_2}  V_{j_2} \right) \right\}, \\
V_{j_2} & = & \int_{\mathcal{M}} E_{\by_{d^{j_2,s+1}}|\mathcal{Y}^{j,s}} \left( p(\x;\mathcal{Y}^{j,s+1}) ( 1-p(\x;\mathcal{Y}^{j,s+1}) ) \right) d\mes(u), \nonumber \\
    p(\x;\mathcal{Y}^{j,s+1}) &=& P(Z(\x) \in T |\by_{d^{j_2,s+1}},\mathcal{Y}^{j,s}). \nonumber
\end{eqnarray}
Here, $\mathcal{Y}^{j,s}=\{\by_{d^{j,s}},\mathcal{Y}^{s-1}\}$ and
$\mathcal{Y}^{j,s+1}=\{\by_{d^{j_2,s+1}},\mathcal{Y}^{j,s}\}$
represent the sets of data variables, and the expectations are
calculated with respect to the conditional densities given data at
that stage.  We solve the first expectation by Monte
Carlo sampling of data $\by_{d^{j,s}}$ from its conditional
distribution. For each of these data samples, the second expectation
is solved using the closed-form expressions for EIBV using the results from Proposition \ref{propo2}, now with conditioning to the first stage data already going into the mean and covariances in Eq. \eqref{eq:krigSUR}.

Even though the strategy looks at two sampling stages, it is only used to find
the current best design. When data are collected at stage $s$, the GP model is
updated as in Eqn. \eqref{eq:krigSUR}, and the mean and (cross) covariance are
used to compute the next design for stage $s+1$, now anticipating what
stage $s+1$ and $s+2$ could offer, and so on.
This look-ahead approach is much more computationally demanding than
the myopic strategy, so for practical implementation onboard the AUV this is reaching the limits allowed for calculation. We conduct some pruning of the least valuable design paths in the evaluation of Eq. \eqref{critLA}, to get a satisfactory computation time below one minute per waypoint. This means that we use the myopic
strategy to rank the three best designs following the first stage alone, and
for each of these preferred designs, we undertake the look-ahead calculations.
